---
title: "stickleback main analysis"
author: "Agnes Olin"
date: "2024-01-04"
output: html_document
---

## Load packages

```{r packages, include=FALSE, message = FALSE}

# packages (R version 4.2.1)
library(broom.mixed) # 0.2.9.4
library(colorspace) # 2.1-0
library(data.table) # 1.14.8
library(DHARMa) # 0.4.6
library(dplyr) # 1.1.3
library(fasterize) # 1.0.4
library(future) # 1.32.0
library(future.apply) # 1.10.0
library(ggnewscale) # 0.4.8
library(ggplot2) # 3.4.1
library(ggpubr) # 0.6.0
library(glmmTMB) # 1.1.7
library(grid) # 4.2.1
library(gstat) # 2.1-1
library(Hmisc) # 5.0-1 
library(INLA) # 22.12.16  
library(lubridate) # 1.9.2 
library(mctest) # 1.3.1
library(MetBrewer) # 0.2.0 
library(MuMIn) # 1.47.5
library(ncdf4) # 1.21
library(ncf) # 1.3-2
library(nlme) # 3.1-157
library(RANN) # 2.6.1
library(raster) # 3.6-23
library(RColorBrewer) # 1.1-3
library(rgdal) # 1.6-4
library(sdmTMB) # 0.3.0
library(seegSDM) # 0.1-9
library(sf) # 1.0-14
library(sjPlot) # 2.8.13
library(visreg) # 2.7.0


# set default figure theme settings
theme_sets = theme(
  text = element_text(family = "sans"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  strip.background = element_rect(fill = "grey90"),
  legend.key.width = unit(1,"cm"))


```



## Load data


```{r load data, warning = FALSE}


## juvenile data ##
# source("help_scripts/JuvenileDataCleanup.R") # script for initial cleanup of juvenile dataset
df_juv = read.csv("data/yngel_final.csv") # juvenile dataset (after cleanup)
df_juv_full = read.csv("data/yngel_final_full.csv") # juvenile dataset (with some additional environmental data)
df_Eklof = read.csv("data/Eklof_detonation_data.csv", sep = ";") # juvenile data from Eklöf et al. 2020 (for Fig 1)

## raster data ##
land5 = raster("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/Landmask5m.TIF") #landmask 5m res
land25 = raster("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/Sweden_NyKustlinje_25m.TIF") # landmask 25m res
SWM = raster("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/swm_10m.TIF") # wave exposure
SWM5 = raster("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/SWM5.TIF") # wave exposure 5m
depth = raster("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/Depth250.TIF") # depth
distance_baseline = raster("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/DistToOffshore_EuclDist_5m_r.TIF") # distance to baseline
baseline = read.csv("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/baseline_points.csv", sep = ";") # baseline in point-format

## BIAS data ##
BIAS = read.csv("data/BIAS_stsp_210209.csv", sep = ";") # BIAS data
recs = read.csv("data/rectangles.csv", sep=";") # info on ICES statistical rectangles


## coastline ##
coast = st_read("//storage.slu.se/home_2$/agin0002/Documents/SLU-Agnes/general_GIS/Europe_coastline/Europe_coastline_poly.shp")
geom_coast = st_geometry(coast)

```


## Clean up juvenile dataset and add environmental data

The juvenile dataset includes 17000+ data points from detonations. Here, the data are cleaned up and some potential drivers added (wave exposure, distance to baseline, year, cormorants, seals, fishing pressure, temperature).

```{r clean up data, include = FALSE, warning = FALSE}

# first add local temperature measurements to the reduced dataset
df_juv$local_SST = df_juv_full$SurfTemp
df_juv$local_BT = df_juv_full$BottomTemp

# remove Finnish data
df_juv = df_juv[!(df_juv$longitude > 20 & df_juv$latitude < 62),]

# add year variable
df_juv$year = year(df_juv$date)

# adding point_id so that points can be related across datasets
df_juv$point_id = 1:nrow(df_juv)

# remove repeated data from 2019
df2019 = df_juv[df_juv$year == 2019, ] # subset
df2019 = df2019[order(df2019$stickleback, decreasing = T), ]
df2019 = df2019[!duplicated(cbind(df2019$latitude, df2019$longitude)),]
df_juv = df_juv[df_juv$year != 2019,]
df_juv = rbind(df_juv, df2019)


# add environmental variables (or load file below)
#df_juv$X = NULL
#source("help_scripts/addSWMandDISTANCE.R") # add SWM and distance to baseline data
#source("help_scripts/addPREDandFISHING.R") # add predation and fisheries data
# write.csv(df_juv, "data/df_juv.csv", row.names = FALSE) # save data with env variables

# load saved data frame with environmental variables
df_juv = read.csv("data/df_juv.csv")

# add in temperature data
#source("help_scripts/temp_processing.R") # process satellite data
#source("help_scripts/addTEMP.R") # adding degree day temperature

# load saved data frame with temperature too 
df_juv = read.csv("data/df_juv_temp.csv")

# temperature comparison
df_juv$local_SST = as.numeric(df_juv$local_SST)
df_juv$local_SST[df_juv$local_SST < 5 | df_juv$local_SST > 35] = NA # remove some odd values

ggplot(df_juv, aes(x = local_SST, y = temp_comp)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_abline(slope = 1, intercept = 0) +
  
  lims(x = c(7,28), y = c(7,28)) +
  labs(x = expression('Local temperature ('*~degree*C*')'), y = expression('Modelled temperature ('*~degree*C*')')) +
  
  theme_bw() +
  theme_sets

ggsave("suppFigures/tempComp.png", width = 10, height = 10, units = "cm")

# correlation test for local vs satellite temperature
cor.test(df_juv$temp_comp, df_juv$local_SST)
sum(!is.na(df_juv$local_SST) & !is.na(df_juv$temp_comp)) # N = 10831

```




## Add BIAS data

Here we add offshore data on stickleback abundances, ie the stickleback that come to the coast to spawn.

```{r add BIAS data, warning = FALSE}

# process data to get values in right unit and format (biomass spawners per km2) and calculate biomass nearby
#source("help_scripts/BIASsmooth.R")
df_juv = read.csv("data/df_BIAS.csv")




```


## Clean up data

Here we subset to relevant time and locations, filter out some sketchy environmental data, and format some response variables.

```{r cleanup inputs, include = FALSE}

## create a spatio-temporal subset ##
df_sub = df_juv[
  df_juv$year > 2000 & # increases began around the year 2000, better data coverage since 2001
    df_juv$latitude <= 60.5, # limit to an area coherent in terms of ecology and management
]

df_sub = df_sub[
  df_sub$distance <= 40000, # remove as in Eklöf et al. 2020
]

## clean up explanatory variables ##
source("help_scripts/data_filtering.R")


# add spatial variables
C = SpatialPoints(
  coords = cbind(df_sub$longitude, df_sub$latitude),
  proj4string = CRS("+proj=longlat +datum=WGS84")
)
C2 = spTransform(
  C,
  "+proj=utm +zone=33 +ellps=GRS80 +units=m +no_defs"
)
XY = as.data.frame(coordinates(C2)); names(XY) = c("X", "Y")
df_sub$X = XY$X; df_sub$Y = XY$Y




```





## Create connectivity metrics

Here we create connectivity metrics for predator spawning habitat connectivity.

### loading habitat objects, prepping and constructing distance dependence

Start by loading the data and making a plot illustrating dispersal distance weighting.

```{r conn_base, warning = FALSE}


### load and create various objects related to habitat patches ###

# here, habitat maps for both limits 3.2 and 3.5 are created
# source("help_scripts/habitatCALC.R")
# this creates the initial rasters habitat32 and habitat35 - which are then modified in QGIS as described below, and final habitat rasters are loaded

# these are polygons based on the habitat maps
# rasters have been polygonised in QGIS
# patches smaller than a hectare have been removed
# patches have also been subset to match area where juvenile data are available
polygons32 = st_read("data/habitat_polygons/polygons32.shp")
polygons35 = st_read("data/habitat_polygons/polygons35.shp")


# centroids for the polygons, also from QGIS
centroids32 = st_read("data/habitat_centroids/centroids32.shp")
centroids35 = st_read("data/habitat_centroids/centroids35.shp")

# after subsetting polygons these have been rasterised to match extent (and also have small nodes removed)
habitat32 = raster("data/raster32.TIF") # habitat raster with log10(SWM) limit 3.2
habitat35 = raster("data/raster35.TIF") # habitat raster with log10(SWM) limit 3.5



#### make plot of dispersal distances ####

source("help_scripts/dispersal_plot.R")



```


### Focal connectivity

Calculating focal connectivity takes a lot of time and is done on an external server using the script connFocal.R (in folder on_gunvor - gunvor is the name of the server).

```{r connectivity_focal, warning = FALSE}

# load calculated values
res_32 = read.csv("data/res32_df.csv")
res_35 = read.csv("data/res35_df.csv")

# fix formats
res_32$conn32 = as.numeric(res_32$conn32)
res_35$conn35 = as.numeric(res_35$conn35)

## merge into df_sub ##
# load id column
res_32$point_id = read.csv("data/conn_foc_id.csv")$point_id
res_35$point_id = read.csv("data/conn_foc_id.csv")$point_id

# merge
df_sub = merge(df_sub, res_32[, c("point_id", "conn32")], by = "point_id", all.x = T)
df_sub = merge(df_sub, res_35[, c("point_id", "conn35")], by = "point_id", all.x = T)


# make histogram of values

maxV = max(c(df_sub$conn32[!is.infinite(df_sub$conn32)], df_sub$conn35[!is.infinite(df_sub$conn35)]), na.rm = T)/10000

p1 = ggplot(data = df_sub, aes(conn35/10000)) +
  geom_histogram(binwidth = 0.1) +
  theme_bw() +
  labs(x = "Connected area (ha)", y = "Count", title = "Limit = 3.5") +
  lims(x = c(0, maxV), y = c(0, 510)) +
  theme_sets 

p2 = ggplot(data = df_sub, aes(conn32/10000)) +
  geom_histogram(binwidth = 0.1) +
  theme_bw() +
  labs(x = "Connected area (ha)", y = "Count", title = "Limit = 3.2") +
  lims(x = c(0, maxV), y = c(0, 510)) +
  theme_sets 


ggarrange(p1, p2,  labels = c("a.", "b."), font.label = list(size = 14, family = "sans", face = "plain"))
ggsave("suppFigures/focalConnHist.jpg", width = 18.5, height = 9, units = "cm")


```





### Network connectivity

Network links are also calculated on an external server using the script conn_network.

```{r connectivity_network, warning = FALSE}


# load calculated values
network_32 = read.csv("data/res32_network.csv")
network_35 = read.csv("data/res35_network.csv")

# extract values for each detonation point
source("help_scripts/extractNetworkMetrics.R")




```



## Sort out repeated samples and fix response variables


```{r cleanup inputs 2, include = FALSE}


## make averages in cases of samples taken at exact same location in exact same year ##

bays = aggregate(bay ~ year + X + Y, data = df_sub, function(x) x[1]) # to be able to add in bay variable later

df_sub = df_sub[, c("year", "X", "Y", "latitude", "pike", "perch", "stickleback", "distance", "BIASmean", "SWM", "fishing", "totalTopPred", "seal", "corm2", "DD", "conn35", "conn32", "area_net_32", "connected_area_net_32", "node_id_32", "area_net_35", "connected_area_net_35", "node_id_35")] # susbet to relevant columns

df_sub = df_sub %>%
  group_by(year, X, Y) %>% 
  summarise_each(funs(mean)) # averages of sample in exact same location in same year
df_sub = as.data.frame(df_sub)

df_sub$bay = bays$bay # add bay variable back in


## fix response variables ##
df_sub$fishPred = round(df_sub$pike+df_sub$perch) # add pike and perch together and round to whole numbers for model
df_sub$stickleback_int = round(df_sub$stickleback) # round stickleback to whole numbers for model
df_sub$RPD = (df_sub$perch + df_sub$pike)/(df_sub$perch + df_sub$pike + df_sub$stickleback) # create dominance ratio


## save final data frame used for analysis ##
write.csv(df_sub, "data/df_sub.csv", row.names = FALSE)



```


## Make maps of predictors and look at spation autocorrelation



```{r predictor maps, include = FALSE}


## look at autocorrelation in predation pressure ##
# source("help_scripts/TopPredFishingAutocorr.R")

# make maps
#source("help_scripts/maps_predictors.R")



```



## Make background plot

Background plot for 

```{r model_prep, include = FALSE}

#source("help_scripts/Fig1.R")




```






## Fit statistical models

Start by making some preliminary plot exploring patterns in the data (structure of response variable, spatial autocorrelation in response and explanatory variables, collinearity).

```{r model_prep, include = FALSE}

#source("help_scripts/pre_model_checks.R")




```


Now - fit the actual models for the different response variables.

```{r models, include = FALSE}


source("help_scripts/models_RPD.R") # models of predator dominance
source("help_scripts/models_pp.R") # model of predator densities
source("help_scripts/models_stick.R") # model of stickleback densities

# combining and saving plots
source("help_scripts/final_plots.R") 


## save comparison of spatio-temporal models ##
# spat_temp_model_comp = cbind(RPD_spat_temp_model_comp,
#                             pp_spat_temp_model_comp,
#                             stick_spat_temp_model_comp)
# 
# tab_df(spat_temp_model_comp,
#        file="result_tables/spat_temp_model_comp.doc")





```


## time series


```{r time series, include = FALSE}


bay_patterns = 
  df_juv %>%
  group_by(year, bay) %>%
  summarise(
    RPD = sum(c(pike, perch))/sum(c(pike, perch, stickleback)),
    N = n()
    
  )

bay_patterns = subset(bay_patterns, !is.nan(RPD) & N >= 5)

bay_sub = subset(bay_patterns, bay %in%   names(table(bay_patterns$bay))[table(bay_patterns$bay) >= 8]  )

bay_sub$col_factor = "mid"
bay_sub$col_factor[bay_sub$RPD <= 0.1] = "stickleback"
bay_sub$col_factor[bay_sub$RPD >= 0.9] = "pred"

bay_sub$year = as.integer(bay_sub$year)

ggplot(bay_sub, aes(year, RPD)) +
  geom_point(aes(colour = col_factor)) +
  scale_colour_manual(values = c("grey", "aquamarine4", "chocolate3")) +
  scale_x_continuous(breaks = seq(1980, 2020, 5)) +
  facet_wrap(~bay, ncol = 5, scales = "free_x") +
  theme_bw(base_size = 7) +
  theme(legend.position = "none",axis.text.x = element_text(angle = -90))

ggsave("suppFigures/time_trends.jpg", width = 18.5, height = 20.0, units = "cm", dpi = 300) 


```


